From Reversible Programs to \\ Univalent Universes and Back

### 2021-05-06 Pi+/paper/univalent.tex
Add all proofs in section 3.2
Fixes #104

\begin{example}[$\BAut$]
  The type of self-equivalences on a type $T$ is defined as $\Aut[T] \defeq T \eqv T$. The subuniverse of types that are
  merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the delooping of $\Aut[T]$, as
  justified by~\cref{lem:loop-deloop}.
\end{example}

\begin{example}
  The univalent subuniverse of types merely equal to $\Bool$ is given by the subtype $P(X) \defeq \SubP{X}{\Bool}$. This
  can be used to give the denotational semantics of a 2-bit reversible
  language~\cite{caretteReversibleProgramsUnivalent2018}.
\end{example}

@@

\begin{example}
  For any $n : \Nat$, we define the subtype of types merely equal to $\Fin[n]$ as $P(X) \defeq \SubP{X}{\Fin[n]}$. This
  gives the univalent subuniverse $\UFin[n] \defeq \Sub{\Fin[n]}$ of finite types with cardinality equal to $n$.
\end{example}

@@

\begin{proof}
  Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
  enough to show that $n \id m$. Since $\Nat$ is a set, this is a proposition, so we can use the induction principle of
  propositional truncation to eliminate to $n \id m$, applying it on $\phi$ and $\psi$ respectively. This gives us the
  equalities $X \id \Fin[n]$ and $X \id \Fin[m]$, which gives us $\Fin[n] \id \Fin[m]$, from which $n \id m$ follows by
  taking the first projection.
\end{proof}

\begin{example}
  The univalent subuniverse of \emph{all finite types} is given by
  \[
    \UFin \defeq \dsum{X:\UU}{\isFin[X]}.
  \]
  We use this as the denotational semantics to interpret our reversible programming language $\PiPlusLang$.
\end{example}

We characterise the path space of subunivalent universes.

@@

\begin{proposition}
  $\UFin$ is a 1-type, that is, a 1-groupoid, or has h-level 3.
\end{proposition}

\begin{lemma}~\label{lem:loop-deloop}
  \[
    \Omega\BAut[T] \eqv \Aut[T]
  \]
\end{lemma}

\begin{proposition}
  For every $n:\Nat$,
  \[
    \Omega\UFin[n] \eqv \Aut[\Fin[n]]
  \]
\end{proposition}

>>>>

\begin{example}[$\BAut$]
  The type of self-equivalences on a type $T$ is defined as $\Aut[T] \defeq T \eqv T$, which is an $\infty$-group. The
  subuniverse of types that are merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the
  classifying space of $\Aut[T]$, as justified by~\cref{lem:loop-deloop}. We write $T_{o} \defeq (T, \trunc{\refl_{T}})$
  for the image of the inclusion of $T$.
\end{example}

\begin{example}
  The univalent subuniverse of types merely equal to $\Bool$ is given by $\BAut[\Bool]$. This can be used to give the
  denotational semantics of a 2-bit reversible language~\cite{caretteReversibleProgramsUnivalent2018}.
\end{example}

@@

\begin{example}
  For any $n : \Nat$, we define the univalent subuniverse of types merely equal to $\Fin[n]$ as $\BAut[\Fin[n]]$. This
  is the universe of finite sets with cardinality equal to $n$. We write $F_{n} \defeq (\Fin[n], \trunc{\refl})$ for the
  image of the inclusion of $\Fin[n]$.
\end{example}

@@

\begin{proof}
  Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
  enough to show that $n \id m$. Since $\Nat$ is a set, this is a proposition, so we can use the induction principle of
  propositional truncation to eliminate to $n \id m$, applying it on $\phi$ and $\psi$ respectively. This gives us the
  equalities $X \id \Fin[n]$ and $X \id \Fin[m]$, which gives us $\Fin[n] \id \Fin[m]$, from which $n \id m$ follows by
  applying the first projection.
\end{proof}

\begin{example}
  The univalent subuniverse of \emph{all finite types} is given by
  \[
    \UFin \defeq \dsum{X:\UU}{\isFin[X]}.
  \]
  We write $F_{n} \defeq (n, \Fin[n], \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$. We use this as the
  denotational semantics to interpret our reversible programming language $\PiPlusLang$.
\end{example}

We characterise the path space of univalent subuniverses.

@@

\begin{proof}
  We need to show that the equality type of $\BAut[T]$ is an $n$-type. Assume $X, Y : \BAut[T]$. Since $\BAut[T]$ is a
  univalent subuniverse, we have $(X \id Y) \eqv (\fst(X) \eqv \fst(Y))$. Note that being an $n$-type is a proposition.
  Since $T$ is an $n$-type, and $\fst(X)$ and $\fst(Y)$ are merely equal to $T$, they're also $n$-types. It follows that
  $\fst(X) \eqv \fst(Y)$ is an $n$-type, and hence $X \id Y$ is an $n$-type.
\end{proof}

\begin{corollary}
  $\UFin$ is a 1-type, that is, a 1-groupoid, or has h-level 3.
\end{corollary}

\begin{lemma}~\label{lem:loop-deloop}
  \[
    \loopspace[\BAut[T],T_{0}] \eqv \Aut[T]
  \]
\end{lemma}

\begin{proof}
  Since $\BAut[T]$ is a univalent universe, it follows that
  \[
    (T_{0} \id_{\BAut[T]} T_{0}) \eqv (\fst(T_{0}) \eqv \fst(T_{0})) \equiv (T \eqv T) \equiv \Aut[T].
  \]
\end{proof}

\begin{corollary}
  For every $n:\Nat$,
  \[
    \loopspace[\UFin[n],\Fin[n]] \eqv \Aut[\Fin[n]]
  \]
\end{corollary}
